Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
bib file included, tweaked a couple code listings (now using verbatimsmall). Bib-related changes to .gitignore.

Computational biologists detect homologies by building 
algorithms which, given an amino-acid query sequence,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{hmmer}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
query sequence element $x_{i}$.
In Haskell, we represent a node as a group of transition probabilities
\begin{verbatim}
data TProbs = TProbs { m_m :: TProb
                     , m_i :: TProb
                     , m_d :: TProb
                     , i_m :: TProb
                     , i_i :: TProb
                     , d_m :: TProb
                     , d_d :: TProb
                     }
\end{verbatim}
and a vector of emission probabilities for the match and insertion states.
\begin{verbatim}
type EProbs = V.Vector logProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{verbatim}

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967}.

@@

To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence.
In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.
The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{durbin} or the HMMER literature~\cite{eddy} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

 simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
SMURFLite~\cite{smurflite}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Beta strands are regions of protein 
structure that ``stick'' to one another.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

@@

Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
strands as ``beads'' which can be slid along a string.
In this analogy, the 
string is the sequence of amino acids of the query protein.
A given placement 
of these beads corresponds to an assignment of particular amino acid residues 
being hydrogen bonded to other particular amino acid residues, and thus MRFy 
computes a probability based on observed frequencies of residues hydrogen 
bonding in beta strands~\cite{betawrappro}.
In between these beta strands, the 
Markov random field is in fact a traditional hidden Markov model, whose optimal 
solution can be easily computed via the Viterbi algorithm.
MRFy performs a 
stochastic search to place these beta strands, based on an objective function 
that combines the probability computed by the Viterbi algorithm with that 
computed from the paired, hydrogen-bonded residues in the beta strands.
If a 
structure has many beta strands, then it also has many miniature hidden Markov 
models which can be solved via the Viterbi algorithm.

@@

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{matt}.

@@

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

>>>>

Computational biologists detect homologies by building 
algorithms which, given an amino-acid query sequence,
find known proteins of similar structure.
When the similar proteins are formed from amino-acid sequences that
are nearly identical to the sequence under test, the homologies can be
detected by  
simple string-matching algorithms.\remark{Can we just drop the
  sentence about string matching?}
When the similar proteins are formed from amino-acid sequences that
are not too far in edit distance from the sequence under test, the homologies can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---the MRFy algorithm---can detect homologies
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.

@@

Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
query sequence element $x_{i}$.
In Haskell, we represent a node as a group of transition probabilities
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb, d_m :: TProb
            , d_d :: TProb}
\end{smallverbatim}
and a vector of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs = V.Vector logProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{smallverbatim}

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967hq}.

@@

To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence.
In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

 simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{Menke:2010ti} and 
SMURFLite~\cite{Daniels:2012}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Beta strands are regions of protein 
structure that ``stick'' to one another.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

@@

Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
strands as ``beads'' which can be slid along a string.
In this analogy, the 
string is the sequence of amino acids of the query protein.
A given placement 
of these beads corresponds to an assignment of particular amino acid residues 
being hydrogen bonded to other particular amino acid residues, and thus MRFy 
computes a probability based on observed frequencies of residues hydrogen 
bonding in beta strands~\cite{Cowen:2002p588}.
In between these beta strands, the 
Markov random field is in fact a traditional hidden Markov model, whose optimal 
solution can be easily computed via the Viterbi algorithm.
MRFy performs a 
stochastic search to place these beta strands, based on an objective function 
that combines the probability computed by the Viterbi algorithm with that 
computed from the paired, hydrogen-bonded residues in the beta strands.
If a 
structure has many beta strands, then it also has many miniature hidden Markov 
models which can be solved via the Viterbi algorithm.

@@

Our previous computational biology projects include an enhancement to a protein 
structural aligner that previously came from our group, Matt~\cite{Menke:2008wu}.

@@

\bibliography{mrfy_experience_report}
